\documentclass{article}
\usepackage{german}
\usepackage[latin1]{inputenc}
\usepackage{a4wide}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{alltt}

\pagestyle{empty}

\begin{document}
\noindent
{\Large \textbf{Exercise}: Converting the Infidels}
\vspace{0.5cm}


\noindent
Three missionaries and three infidels have to cross a river. Their boat 
only provides room for two passengers. The infidels are also cannibals
and are quite hungry.  If there are fewer missionaries than infidels on a
shore, then the missionaries have a problem.  Your task is to create a shedule
that can be used to cross the river.  The constraint is that the 
missionaries must not be consumed by the infidels.

You will find a frame for this program on my web page at
\\[0.1cm]
\hspace*{0.8cm}      
\texttt{http://www.ba-stuttgart.de/\symbol{126}stroetma/SETL2/missionare-frame.stl} 
\\[0.1cm]
Your task is to complete this frame to a running program solving the puzzle.
\begin{enumerate}
\item Define a procedure $\texttt{problem}(M, I)$ where $M$ is the number of missionaries
      on the left shore and $I$ is the number of infidels on the left shore.
      The procedure should return \texttt{true} if and only if there is a problem
      on either the left or the right shore.

      (Line 71 in \texttt{missionary-frame.stl})
\item Define a set $P$ of points.  Every point has to describe a situation.  Represent a
      situation as a triple $\langle M, I, B \rangle$,
      where  $M$ is the number of missionaries on the left shore, $I$ is the number of
      infidels on the left shore and $B$ is the number of Boats on the left shore.
      The set $P$ shall contain only those states where there is no problem for the
      missionaries.

      (Line 17 in \texttt{missionary-frame.stl})
\item Define a relation $R_{1}$ on the set $P$.
      A pair
      \[ \bigl\langle \langle M_1, I_1, B_1 \rangle, \langle M_2, I_2, B_2 \rangle \bigr\rangle \]
      should be a member of
      $R_{1}$ iff the state $\langle M_1, I_1, B_1\rangle$ can be transformed into the state
      $\langle M_2, I_2, B_2\rangle$
      by crossing the boat from the left shore to the right shore.
      Additionally, the missionaries should not have a problem after the crossing.

      (Line 28 in \texttt{missionary-frame.stl})
\item In the same way, define the relation $R_{2}$ as the relation describing the
      crossings from the right shore to the left shore.    

  
      (Line 43 in \texttt{missionary-frame.stl}.)
\item Define the start state and the goal state.

      (Line 61 und 63 in \texttt{missionary-frame.stl}.)

      \textbf{Remark}:  The procedure  \texttt{reachable} is already a part of the file
       \texttt{missionary-frame.stl}.  This procedure is called in line 65.  Furthermore,
       the file contains a number of predefined procedures that are used to output
       intermediary results.  If your program does not work immediately, then you
       should better check these results.
\end{enumerate}
\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
