\documentclass{article}
\usepackage{german}
\usepackage[latin1]{inputenc}
\usepackage{a4wide}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{alltt}

\pagestyle{empty}

\begin{document}
\noindent
{\Large \textbf{Exercise}: The Wine Trader}
\vspace{1.0cm}



\noindent
A wine trader has a barrel that is full of wine and two empty buckets.
The first bucket has a volume of 3 litres and the second bucket has a volume of 5 litres.
Now the trader wants to fill the second bucket with exactly 4 litres of wine.  
In order to do this, the following operations are permitted:
\begin{enumerate}
\item A bucket can be filled completely with wine from the barrel.
\item A bucket can be emptied completely into the barrel.
\item The content of one bucket $A$ can be filled into another bucket $B$.
      There are two cases:
      \begin{enumerate}
      \item If there is enough room in the bucket $B$, the 
            bucket $A$ is emptied completely.
      \item If there is not enough room in the bucket $B$,  the  bucket $B$
            is filled completely and the rest of the wine remains in the bucket $A$.
      \end{enumerate}
\end{enumerate}
Your task is to implement a \textsc{Setl2} program that shows how the first bucket can be
filled with exactly 4 litres of wine.  
 I have prepared a file that can be found at:
\\[0.2cm]
\hspace*{1.3cm}
\texttt{http://www.ba-stuttgart.de/\symbol{126}stroetma/SETL2/wine-trader-frame.stl} 
\\[0.2cm]
Your task is to fill the missing gaps in this program.

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
